---
title: "Introduction to Macroeconomic Analysis Through Financial Programming"
author: "Bryce Quillin"
date: "https://github.com/bquillin12/lecture_external_accounts.git"
output:
  beamer_presentation: default
  slidy_presentation: default
  ioslides_presentation: default
subtitle: Focus on External and Fiscal Accounts
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = FALSE)
```

##  General framework
* Real economy (output, price level, exchange rates)
* _Government sector (fiscal revenues and expenditures)_
* Monetary and financial sector 
* _External sector (BoP and IIP)_

__Modeling requires these four sectors to be consistent with one another__

## A case study: Thailand in 1997

```{r gdp, echo=FALSE, message=FALSE, warning=FALSE}
knitr::opts_chunk$set(echo = FALSE, warning = FALSE)

library(reshape2)
library(ggplot2)
thailand_gdp <- read.csv("thailand_gdp.csv", sep = ",", header = T, fileEncoding="UTF-8-BOM")
thailand_gdp <- melt(thailand_gdp, id.vars = "Year")
ggplot(thailand_gdp) +
 aes(x = Year, y = value, fill = variable, colour = variable) +
 geom_line(size = 1.3) +
 scale_fill_hue() +
 scale_color_hue() +
 labs(title = "GDP Growth (y/y, %), 1990-98", caption = "Source: International Monetary Fund") +
 theme_minimal() + theme(plot.title = element_text(hjust = 0.5), plot.caption = element_text(hjust = 0.0),
        legend.title = element_blank()) + xlab("") + ylab("")
```

## The fiscal deficit deteriorated

```{r structural_balance, echo=FALSE, message=FALSE, warning=FALSE}
knitr::opts_chunk$set(echo = FALSE, warning = FALSE)

library(ggplot2)
structural_balance <- read.csv("structural_balance.csv", sep = ",", header = T, fileEncoding="UTF-8-BOM")
ggplot(structural_balance) +
  aes(x = year, y = balance, color = "red") +
  geom_line(size = 1.3) +
  scale_fill_hue() +
 scale_color_hue() +
 labs(title = "Thailand: Structural Fiscal Balance (% Potential GDP), 1995-98", caption = "Source: International Monetary Fund") +
  theme_minimal() + theme(plot.title = element_text(hjust = 0.5), plot.caption = element_text(hjust = 0.0)) +
  theme(legend.position = "none") +
  xlab("") + ylab("")
  
```

## The current account improved as demand collapsed

```{r cab, echo=FALSE, message=FALSE, warning=FALSE}
knitr::opts_chunk$set(echo = FALSE, warning = FALSE)

library(ggplot2)
cab <- read.csv("cab.csv", sep = ",", header = T, fileEncoding="UTF-8-BOM")
ggplot(cab) +
  aes(x = year, y = balance, color = "red") +
  geom_line(size = 1.3) +
  scale_fill_hue() +
 scale_color_hue() +
 labs(title = "Thailand: Current Account Balance (% GDP), 1990-98", caption = "Source: International Monetary Fund") +
  theme_minimal() + theme(plot.title = element_text(hjust = 0.5), plot.caption = element_text(hjust = 0.0)) +
  theme(legend.position = "none") +
  xlab("") + ylab("") +
  scale_x_continuous(name = " ", breaks = c(1990, 1991, 1992, 1993, 1994, 1995, 1996, 1997, 1998))
  
```

## Reserves disappeared fast

```{r reserves, echo=FALSE, message=FALSE, warning=FALSE}
knitr::opts_chunk$set(echo = FALSE, warning = FALSE)

year <- c(1995, 1996, 1997, 1998, 1999)
reserves_amount <- c(36.9, 38.6, 27.0, 29.5, 34.8)
reserves_amount_df <- data.frame(year, reserves_amount)

library(ggplot2)
ggplot(reserves_amount_df) +
   aes(x = year, y = reserves_amount, color = "red") +
  geom_line(size = 1.3) +
  scale_fill_hue() +
 scale_color_hue() +
 labs(title = "Thailand: Total Reserves (incl. gold, current USD), 1995-99", caption = "Source: International Monetary Fund") +
  theme_minimal() + theme(plot.title = element_text(hjust = 0.5), plot.caption = element_text(hjust = 0.0)) +
  theme(legend.position = "none") +
  xlab("") + ylab("")
```

## External debt stocks exploded

```{r external_debt, echo = F, message = F, warning = F}
knitr::opts_chunk$set(echo = FALSE, warning = FALSE)

year <- c(1995, 1996, 1997, 1998, 1999)
ext_debt <- c(60.2, 63.3, 75.3, 96.0, 78.9)
external_debt <- data.frame(year, ext_debt)

library(ggplot2)
ggplot(external_debt, aes(x = year, y = ext_debt)) + geom_bar(stat = "identity", color = "red", fill = "red") +
  labs(title = "Thailand: External debt (% GDP), 1995-99", caption = "Source: World Bank.") +
  theme_minimal() + theme(plot.title = element_text(hjust = 0.5), plot.caption = element_text(hjust = 0.0)) +
  xlab("") + ylab("")
  
  

```

## Large stocks of debt were on short maturities

```{r st_debt, echo = F, message = F, warning = F}
knitr::opts_chunk$set(echo = FALSE, warning = FALSE)

year <- c(1995, 1996, 1997, 1998, 1999)
st_debt <- c(119, 123, 140, 100, 67)
st_debt_df <- data.frame(year, st_debt)

library(ggplot2)
ggplot(st_debt_df, aes(x = year, y = st_debt)) + geom_bar(stat = "identity", color = "red", fill = "red") +
  labs(title = "Thailand: Short-Term Debt (% total reserves), 1995-99", subtitle = "Claims between 1-2 years", caption = "Source: World Bank.") +
  theme_minimal() + theme(plot.title = element_text(hjust = 0.5), plot.caption = element_text(hjust = 0.0)) +
  xlab("") + ylab("")

```


## External sector
* Free trade a rare area of consensus among economists
* International trade allows global production to be organized more efficiently: international division of labor
* Breaks the identity that domestic consumption and investment = domestic production
* Spending in excess of domestic production can be financed by accumulating external liabilities
* Excess of production over spending results in accumulation of external wealth

## External sector and international political economy
* External imbalance tell you a lot about a country
  + Its stage of development
  + The adquacy of its policies (macroeconomic as well as structural)
  + A major source of economic and financial crises, perhaps with politic origins ans always with political consequences
* Cross-national external imbalances a source of international conflict
  + Debates about excessive surpluses or deficits
  + Long standing debate on "who adjusts?"
* Involves the contribution of international institutions (e.g., IMF)

## Balance of payments - some basics
* Records transactions of goods, services, and assets between residents and foreign residents in two separate accounts:
    + Current account
    + Capital and financial account
* Transactions recorded in a major international currency, normally USD
* Complied on accrual accounting basis
* Constructed with double entry accounting system: + is a credit and - a debit.

## BOP - Current account
* Goods - Trade balance = exports - imports of goods
* Services
    + Transporation/travel
    + Government services
    +Other services
* Income
    + Compensation of employees
    + Investment income
* Current transfers

__The CA balance is often short hand for a country's external position__

## BOP Capital account
  * Capital account: 
    + Capital transfers 
    + Acquisition/disposal of non-produced, nonfinancial assets

* Financial account 
    + Direct investment (FDI), net
    + Portfolio investment, net
    + Other investments, net
    + Loans, trade credits
    + Reserve assets - finance imbalances elsewhere in BOP (important to understand)
    
## Determinants of CAB: savings and investment - absorption approach
* CAB = Gross national disposable income (GNDI) - Absorption (A)
* CA deficit occurs when  a country spends beyond its means or absorbs more than it produces
* Thus, reducing CA deficit requires an increase in income and/or reduction in absorption
* Increasing output (and hence income): 
    + Short-term: requires unused capacity
    + Medium-term: increased production capacity through investment, labor force participation, and structural policies
* Reducing absorption achieved by contracting final consumption or investment

## Determinants of CAB: savings and investment - SI approach
* Savings-investment balance approach
    + CAB = Savings - Investment
* In this view, CAB reflects country's use of foreign savings
* Excess of investment over savings must be covered by foreign saving
* CA deficit thus reduced by increasing savings and/or reducing investment

## Determiants of CAB: capital flows and reserves
* The CAB identities imply:
    + CAB + Net Capital Inflow (FI) - Change in Reserves (delta RES) = 0
* CA deficit can be maintained only as long as capital inflows persist and/or net official international reserves decline
* Large and persistent CA deficits and increasing net international indebtedness raise questions over sustainability

## Are current account deficits or surpluses bad? (1)
* In principle, deficits and surpluses are normal and may be related to stages of development

* Lower income countries might run CA deficits as part of economic development: there are profitable domestic investment opportunities but insufficient finance (capital should flow "downhill")

* CA deficits may be optimal response to temporary negative shock

* Aging economies may run CA surpluses as foreign assets are accumulated to fund retirement consumption

* CA surpluses may reflect attempt to "insure" against financial crises or cyclical downturns in commodities prices

## Are current account deficits or surpluses bad? (2)

* On the other hand, CA surpluses or deficits may reflect broader
problems in macroeconomic or structural policies

* For example, excessive domestic demand pressures may result from
loose fiscal policy, translating into higher imports with no
corresponding increase in exports

* Or expectations for excessively tight monetary policy or weak labor
markets may encourage high household savings rates and surpress
demand for imported goods

## Indicators of external vulnerability

* In countries without or limited capital markets access, CA deficits are
financed through changes in gross foreign reserves. A risk indicator
for these countries is the
stock of international reserves to monthly
import bill.
The standard is normally that three months is sufficient

* For those with capital markets access, CA deficits can be financed
through foreign borrowing, portfolio inflows, or FDI. The composition
of financing matters, some specific metrics include:

+ Ratio of short term debt to reserves
+ Exchange rate behavior
+ Interest rate spreads against a benchmark, such as U.S. government bonds
+ Is country vulnerable to financing disruptions

## Current account and the real exchange rate
* CA position closely linked to the level of the real exchange rate ( =
price of domestic consumption basket vs. foreign consumption
basket)

*  A real exchange rate appreciation occurs when domestically produced
goods become more expensive than goods produced abroad

* As labor is the largest component in most sectors, labor costs are a
big determinant of domestic inflation and unit labor costs are another
basis to measure the real exchange rate

* Normally a real exchange rate appreciation will lead to a deterioration
in the CA

## Now let's look at another case study, Turkey, where growth has slowed..

```{r turkey, echo = F, message = F, warning = F}
knitr::opts_chunk$set(echo = FALSE, warning = FALSE)

year <- c(2017, 2018, 2019, 2020, 2021)

turkey_gdp_growth <- c(7.5, 2.8, 0.2, 3.0, 3.0)
turkey_cab <- c(-5.6, -3.5, -0.1, -0.6, -1.3)
turkey_structural <- c(-1.6, -2.5, -2.9, -1.9, -1.8)
turkey_external_debt <- c(53.4, 57.6, 61.3, 55.7, 50.7)

turkey_gdp_df <- data.frame(year, turkey_gdp_growth)
turkey_cab_df <- data.frame(year, turkey_cab)
turkey_structural_df <- data.frame(year, turkey_structural)
turkey_external_debt_df <- data.frame(year, turkey_external_debt)

library(ggplot2)
library(ggpubr)

turkey_gdp_plot <- ggplot(turkey_gdp_df, aes(x = year, y = turkey_gdp_growth)) + geom_line(color = "red") +
  labs(title = "GDP Growth (y/y, %)") +  theme_minimal() + theme(plot.title = element_text(hjust = 0.5), plot.caption = element_text(hjust = 0.0)) + xlab("") + ylab("") 

turkey_cab_plot <- ggplot(turkey_cab_df, aes(x = year, y = turkey_cab)) + geom_line(color = "red") +
  labs(title = "Current Account Balance (% GDP)") +  theme_minimal() + theme(plot.title = element_text(hjust = 0.5), plot.caption = element_text(hjust = 0.0)) + xlab("") + ylab("") 

turkey_structural_plot <- ggplot(turkey_structural_df, aes(x = year, y = turkey_structural)) + geom_line(color = "red") + labs(title = "Structural Primary Balance (% potential output)", caption = "Source: International Monetary Fund.") +  
  theme_minimal() + theme(plot.title = element_text(hjust = 0.5), plot.caption = element_text(hjust = 0.0)) + xlab("") + ylab("") 

turkey_debt_plot <- ggplot(turkey_external_debt_df, aes(x = year, y = turkey_external_debt)) + geom_line(color = "red") + labs(title = "Gross External Debt (% GDP)") +  theme_minimal() + theme(plot.title = element_text(hjust = 0.5), plot.caption = element_text(hjust = 0.0)) + xlab("") + ylab("") 

ggarrange(turkey_gdp_plot, turkey_cab_plot, turkey_structural_plot, turkey_debt_plot, ncol = 2, nrow = 2)


```














